Trait pleiotropy of regulatory variants increases with the variety of bound transcription factors, regulated genes and active tissues

### 2022-08-27 ms/ms01_titlepage.tex
ENH intro and discussion

\title{Phenotype pleiotropy of regulatory variants increases with the variety of bound transcription factors, regulated genes and active tissues}
%\shorttitle{Pleiotropy and gene regulatory variants}

@@

Pleiotropic genetic variants affecting several phenotypes are relatively common in the genome as shown by the increasing number of genome-wide association studies (GWAS).
%
Frequent variants are often located in non-coding regions and are likely gene regulatory variants.
%
In this work, I investigate the link between gene regulation and phenotype pleiotropy for frequent variants.

% Verify numbers
First I have computed the colocalization of 413 GWAS studies and 127 eQTL studies and have observed 143,119 colocalized pairs with a probability above 0.8.
%
I have categorized the GWAS phenotypes into 96 phenotypes and categories to define pleiotropic as belonging at least to two of these categories.
%
Pleiotropic variants are associated to more eQTL genes (egenes) and more eQTL tissues (etissues) but egenes do also correlate with more phenotypes.
%
These observations could be partially explained, because pleitropic variants are enriched in splicing and missense variants and bind more transcription factors.
%
These variants with the GWAS phenotypes, egenes and etissues can be explored in a public website.

>>>>

\title{Trait pleiotropy of regulatory variants increases with the variety of bound transcription factors, regulated genes and active tissues}
%\shorttitle{Pleiotropy and gene regulatory variants}

@@

Pleiotropic genetic variants affecting several traits are relatively common in the genome as shown by the increasing number of genome-wide association studies (GWAS).
%
Frequent variants are often located in non-coding regions and are likely gene regulatory variants.
%
In this work, I investigate the link between gene regulation and trait pleiotropy for frequent variants.

% Verify numbers
First I have computed the colocalization of 413 GWAS studies and 127 eQTL studies and have observed 143,119 colocalized pairs with a probability above 0.8.
%
I have categorized the GWAS traits into 96 traits and categories to define pleiotropic as belonging at least to two of these categories.
%
Pleiotropic variants are associated to more eQTL genes (egenes) and more eQTL tissues (etissues) but egenes do also correlate with more traits.
%
These observations could be partially explained, because pleitropic variants are enriched in splicing and missense variants and bind more transcription factors.
%
These variants with the GWAS traits, egenes and etissues can be explored in a public website.

### 2022-08-27 ms/ms02_intro.tex
ENH intro and discussion

Genome-wide association studies (GWAS) evaluate the impact of frequent variants on a trait in the population.
%
The number of GWAS has rapidly increased since the first GWAS in 2010 and has result in comprehensive databases such as the NHGRI-EBI GWAS Catalog \citep{2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71673 variant-trait associations from 3567 publications \citep{2018.Parkinson.Buniello}.

@@

Expression QTLs (eQTLs) are variants that change the expression of a given gene in a tissue or cell type.
%
eQTLs have some of the same shortcommings as GWAS variants such as inability to identify the causal variant.
%
On the other hand, eQTLs provide information about the target genes and tissues.

A large number of GWAS variants are found in non-conding regions.
To priori

Sentence describing EBI eQTL

In this work, I investigate the molecular basis of pleiotropy of gene regulatory variants.

>>>>

Genome-wide association studies (GWAS) allow to define the impact of genetic variants on a trait in the population.
%
The number of GWAS has rapidly increased since the first GWAS in 2010 and has result in comprehensive databases such as the NHGRI-EBI GWAS Catalog \citep{2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71673 variant-trait associations from 3567 publications \citep{2018.Parkinson.Buniello}.

@@

Expression QTLs (eQTLs) are variants that change the expression of a given gene in a tissue or cell type.
%
eQTLs share some of the same shortcommings as GWAS variants such as inability to identify the causal variant.
%
Also the distribution of eQTLs and GWAS variants in the genome have been suggested to be different because of negative selection.

Despite these previous disadvantages eQTLs are still important data sets to interprete GWAS variants, because they provide gene regulatory information about the eQTL genes (egenes) and tissues (etissues) (REF).
%
In addition colocalization anlaysis can pinpoint more precisely the causal variants that either data set independently (REF).

The large number of GWAS has allowed to discover that many genetic variants are involved in different traits.
%
TODO explain Watanabe.
%
Because GWAS variants affect mostly non-coding regions, it makes sense to investigate trait pleiotropy in terms of non-coding gene regulatory functions.

For these reasons, I have taken advantage of two well formatted eQTL and GWAS data sets to develop a global colocalization analysis that includes 127 eQTL studies and 413 GWAS.
%
I have used these colocalization results to investigate the gene regulatory properties of pleiotropic variants.
%
This global colocalization analysis is also useful to help interpret GWAS variants.

### 2022-08-27 ms/ms03_results.tex
ENH intro and discussion

Public GWAS summary statistics were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular phenotypes such as proteome or methylome.
%	
The second criterion was to include only the European population, because most samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological conditions and exclude environmental phenotypes such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS studies with at least 10000 subjects, 2000 controls and 2000 cases (Supplementary table 2).
%
These filters resulted in 413 GWAS (Supplementary table 2).

@@

I manually assigned the 413 GWAS to 96 categories to aggregate identical or similar phenotypes (Supplementary table 2).
%
In the EBI eQTL Catalogue, immune cell types are annotated to great detail while other tissues are annotated at lower resolution.
%
Therefore I manually defined 36 categories to annotate eQTL biological samples at equivalent resolution (Supplementary table 1).

These category annotations were used to investigate the variant pleiotropy for GWAS phenotypes.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
Pleiotropic variants aggregate in some cytobands such as 3q23, 5q31.1, 11q13.5, 12q24.12 and 15q26.1 (Table \ref{tab:pleitropic_variants} and Supplementary table 3).

@@

%Variants of the 3q23 cytoband are involved in allergy, cancer and height phenotypes (Supplementary table 3).
%%
%Interestingly all these variants target the egene ZBTB3.
%%
%ZBTB3 is a transcription factor that controls pro-inflammatory factors such as IRF5 (10.1016/j.jaut.2016.08.001).
%%
%These variants are active in tissues such as blood, esophagus, immune cell, LCLs and muscle (Supplementary table 3).

@@

The most pleiotropic region is 12:111,395,984-111,645,358 in cytoband 12q24.12 that we discussed above (Table \ref{tab:pleiotropic_regions}).
%
The largest region is 11:13,260,511-17,396,930 in cytoband 11p15.2 with a length of 4,136,419 bp and variants involved in cardiovascular and hypertension phenotypes (Supplementary table 5).
%
The second largest region is 2:187,235,912-191,066,738 in cytoband 2q32.2 with a length of 3,830,826 bp and variants involved in cardiovascular, hypertension and autoimmune phenotypes  (Supplementary table 5).

@@

\subsection*{How specific are variants to GWAS phenotypes, egenes and etissues?}

Next I explored the phenotype, egene and etissue specificity of eQTL/GWAS variants.
%
83\% of variants are involved in one GWAS category, 14\% are involved in 2 categories and the remaining variants with 3 or more phenotype categories make 3\% or less (Fig. \ref{fig:hist_gwas_egene_etissue}(a).
%
Regarding egenes, half of variants modulate one egene, and the other half modulate two or more egenes (Fig. \ref{fig:hist_gwas_egene_etissue}(b).
%
In the case of etissues, only 40\% of variants are specific to a single tissue while the other 60\% are active in two or more tissues (Fig. \ref{fig:hist_gwas_egene_etissue}(c).
%
In conclusion, most colocalized eQTL/GWAS variants are specific to one GWAS phenotype. By contrast, half of them only module one specific egene, and less than half are active in a single tissue.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{How do phenotype frequencies relates to eQTL and etissue frequency at the loci level?}

In Figure \ref{fig:region_gwas_egenes_tissues}, I have plotted the number of GWAS categories, egenes and eTissues for regulatory variants in three pleiotropic regions.
%
All three regions are very pleiotropic with very different associated phenotype categories such as cancer, cardiovascular and autoimmune diseases.
%
However these plots show that the relationship between the number of phenotypes, egenes and etissues can be very different.
%
The SLC22A5 locus in region 5:132,239,645-132,497,907 (5q31.1) has a low number of egenes but a high number of etissues.
%
By contrast, the MHC locus in region 6:31,034,839-32,478,149 (6p21.33) has a high number of both egenes and etissues.
%
Finally, the ATXN2 locus in region 12:111,395,984-111,645,358 (12q24.12) shows the highest phenotype pleiotropy but the number of egenes and etissues is very low.
%
In summary, phenotype pleiotropy can arise from different situations: few egenes and etissues, many egenes and etissues or a mix of both.

>>>>

Public GWAS summary statistics were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS studies with at least 10000 subjects, 2000 controls and 2000 cases (Supplementary table 2).
%
These filters resulted in 413 GWAS (Supplementary table 2).

@@

I manually assigned the 413 GWAS to 96 categories to aggregate identical or similar traits (Supplementary table 2).
%
In the EBI eQTL Catalogue, immune cell types are annotated to great detail while other tissues are annotated at lower resolution.
%
Therefore I manually defined 36 categories to annotate eQTL biological samples at equivalent resolution (Supplementary table 1).

These category annotations were used to investigate the variant pleiotropy for GWAS traits.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
Pleiotropic variants aggregate in some cytobands such as 3q23, 5q31.1, 11q13.5, 12q24.12 and 15q26.1 (Table \ref{tab:pleitropic_variants} and Supplementary table 3).

@@

%Variants of the 3q23 cytoband are involved in allergy, cancer and height traits (Supplementary table 3).
%%
%Interestingly all these variants target the egene ZBTB3.
%%
%ZBTB3 is a transcription factor that controls pro-inflammatory factors such as IRF5 (10.1016/j.jaut.2016.08.001).
%%
%These variants are active in tissues such as blood, esophagus, immune cell, LCLs and muscle (Supplementary table 3).

@@

The most pleiotropic region is 12:111,395,984-111,645,358 in cytoband 12q24.12 that we discussed above (Table \ref{tab:pleiotropic_regions}).
%
The largest region is 11:13,260,511-17,396,930 in cytoband 11p15.2 with a length of 4,136,419 bp and variants involved in cardiovascular and hypertension traits (Supplementary table 5).
%
The second largest region is 2:187,235,912-191,066,738 in cytoband 2q32.2 with a length of 3,830,826 bp and variants involved in cardiovascular, hypertension and autoimmune traits  (Supplementary table 5).

@@

\subsection*{How specific are variants to GWAS traits, egenes and etissues?}

Next I explored the trait, egene and etissue specificity of eQTL/GWAS variants.
%
83\% of variants are involved in one GWAS category, 14\% are involved in 2 categories and the remaining variants with 3 or more trait categories make 3\% or less (Fig. \ref{fig:hist_gwas_egene_etissue}(a).
%
Regarding egenes, half of variants modulate one egene, and the other half modulate two or more egenes (Fig. \ref{fig:hist_gwas_egene_etissue}(b).
%
In the case of etissues, only 40\% of variants are specific to a single tissue while the other 60\% are active in two or more tissues (Fig. \ref{fig:hist_gwas_egene_etissue}(c).
%
In conclusion, most colocalized eQTL/GWAS variants are specific to one GWAS trait. By contrast, half of them only module one specific egene, and less than half are active in a single tissue.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{How do trait frequencies relates to eQTL and etissue frequency at the loci level?}

In Figure \ref{fig:region_gwas_egenes_tissues}, I have plotted the number of GWAS categories, egenes and eTissues for regulatory variants in three pleiotropic regions.
%
All three regions are very pleiotropic with very different associated trait categories such as cancer, cardiovascular and autoimmune diseases.
%
However these plots show that the relationship between the number of traits, egenes and etissues can be very different.
%
The SLC22A5 locus in region 5:132,239,645-132,497,907 (5q31.1) has a low number of egenes but a high number of etissues.
%
By contrast, the MHC locus in region 6:31,034,839-32,478,149 (6p21.33) has a high number of both egenes and etissues.
%
Finally, the ATXN2 locus in region 12:111,395,984-111,645,358 (12q24.12) shows the highest trait pleiotropy but the number of egenes and etissues is very low.
%
In summary, trait pleiotropy can arise from different situations: few egenes and etissues, many egenes and etissues or a mix of both.

### 2022-08-27 ms/ms04_discussion.tex
ENH intro and discussion

In this work, I present a colocalization pipeline for eQTLs and GWAS variants that I have applied to a large number of 413 GWAS and 127 eQTL studies.
%
For this purpose, I have taken advantage of two public resources of GWAS and eQTL studies with summary statistics to compute a large number of colocalizations between eQTLs and GWAS variants.
%
The results of this pipeline provides three types of insights.
%
First these results allow to investigate the molecular characteristics of pleiotropic variants with a gene regulatory effect.
%
Second, the colocalized loci provide predictions of causal variants in given loci of GWAS.
%
Third, egenes and etissues of colocalized eQTL variants provide potential predictions of the molecular mechanism of many GWAs traits.

% First insight: pleiotropy %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For a posterior probability PP.H4.abf$\ge$0.8 of common causal eQTL and GWAS variants, we found 100,526 loci.
%
If I focus in causal variants with PP.H4.abf$\ge$0.8 and SNP.PP.H4$\ge$0.5, then I find 4684 variants.
%
Our analysis suggests that these pleiotropic variants are under the control of more transcription factors and regulate more egenes in more etissues.

% Second insight: causal variants %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Third insight: mechanistic information for GWAS variants %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

These variants are under control of more transcription factors

%\subsection*{What is the function of the variants in the most pleiotropic regions?}
%
%% 12q24.12
%The cytoband 12q24.12 has five variants rs3184504, rs653178, rs7310615, rs597808, rs11065979, that are known in the literature to be associated to a wide palette of diseases.
%This locus has been used to explain links between inflammaiton and hypertension (10.1097/MNH.0000000000000196), diabetes and autoimmunity (10.4239/wjd.v5.i3.316).
%%
%Our analysis add up that the variants in this cytoband control important genes such as ALDH2, ATXN2, MAPKAPK5, SH2B3 and TMEM116 in several tissues such as adipose, artery, blood, brain, digestive, skin, heart and skin tissue as well as immune cells.
%
%% 6p21.3
%% TODO
%
%%15q26.1
%The next pleiotropic locus is in cytoband 15q26.1 involved in cardiovascular, hypertension and psyciatric diseases with variants rs2071382, rs8039305 and rs6224.
%%
%eQTLs in cytoband 15q26.1 control genes like Furin, oncogene Fes and UNC45A involved in Osteootohepatoenteric syndrome (OMIM DB).
%%
%eQTLs are active in a very large number of cells including blood, artery, brain and immune cells (See supplementary table).
%
%% 1p13.3
%The locus in cytoband 1p13.3 is mainly related to cardiovascular diseases.
%Interestingly, these variants are active in many tissues (Supp table h4_annotated.ods).
%
%
%
%The variant rs2107595 is in locus 7p21.1 downstream of the histone deacetylase 9 (HDAC9) is cardiovascular. Interestingly, it controls expression of lncRNA XXX that is not known in the context of cardiovascular diseases.
%
%The variant rs11236797 and the two others in locus 11q13.5 is mainly autoimmune.
%All three regulate the gene EMSY-DT and are active induced pluripotent cells, LCDs and skin.
%
%Another pleiotropic variants rs301806
%rs301807
%rs301802
%rs301817
%rs159963
%rs301816
%in the cytoband 1p36.23 that are involved in allergy, cardiovascular and hypertension.
%This variants control the gene RERE, which is a nuclear receptor corregulator and is involved in neurodevelopmental disorders.\\

>>>>

In this article, I present the results of a colocalization pipeline for eQTLs and GWAS variants that I have applied to a large number of 413 GWAS and 127 eQTL studies.
%
I have taken advantage of two public resources of GWAS and eQTL studies with summary statistics to compute a large number of colocalizations between eQTLs and GWAS variants.
%
These results are provided as a downloadable and an interactive table that can be used to explore variants with the GWAS traits, etissues and egenes.
%
For a posterior probability PP.H4.abf$\ge$0.8 of common causal eQTL and GWAS variants, we found 100,526 loci.
%
If I focus on causal variants with PP.H4.abf$\ge$0.8 and SNP.PP.H4$\ge$0.5, then I find 4684 variants.

This interactive table can be used for different applications.
%
For instance, egenes and etissues of colocalized eQTL variants provide potential predictions of the molecular mechanism of GWAs traits.
%
Colocalized loci provide predictions of potential causal variants in a given loci of GWAS.
%
The variant rs2107595 is in locus 7p21.1 downstream of the histone deacetylase 9 (HDAC9) is cardiovascular. Interestingly, it controls expression of lncRNA XXX that is not known in the context of cardiovascular diseases.

I have extensively compared molecular properties of pleiotropic variants involved in several GWAS traits.
%
Our analysis provides a number of hints about pleiotropic variants and regions.
%
I have found that these pleiotropic variants are under the control of more transcription factors and regulate more egenes in more etissues.
%
By contrast, the eQTLs of pleiotropic variants show lower beta and less significant p-values (Fig. TODO).
%
This can make sense, because these variants target a larger number of egenes and etissues, and larger effects are maybe negatively selected.
%
It has been recently observed that eQTLs are negatively selected from essential genes (Ref bioarxiv TODO).
%
On the other hand, the effect on GWAS traits is more significant suggesting that the additive effect of several egenes and etissues results in stronger effects at the level of the GWAS trait (Fig TODO).

At the molecular level, I have found that the immune system might be partly involved in pleiotropy.
%
Immune related gene ontology is significantly found in pleiotropic egenes (Fig 1) and etissues related to immune cells are overrepresented in pleiotropic loci (Fig 2).
%
The most pleiotropic locus 12q24.12 includes the SH2B3 gene that regulates cytokine signaling.
%
This locus has been used to explain links between inflammation and hypertension (10.1097/MNH.0000000000000196), diabetes and autoimmunity (10.4239/wjd.v5.i3.316).
%
There is also the 5q31.1 locus that includes important genes of the immune system such as IRF1 and IL4 and the MHC locus that was previously discussed (REF watanaba discussion TODO).

%% 6p21.3
%% TODO
%
%%15q26.1
%The next pleiotropic locus is in cytoband 15q26.1 involved in cardiovascular, hypertension and psyciatric diseases with variants rs2071382, rs8039305 and rs6224.
%%
%eQTLs in cytoband 15q26.1 control genes like Furin, oncogene Fes and UNC45A involved in Osteootohepatoenteric syndrome (OMIM DB).
%%
%eQTLs are active in a very large number of cells including blood, artery, brain and immune cells (See supplementary table).
%
%% 1p13.3
%The locus in cytoband 1p13.3 is mainly related to cardiovascular diseases.
%Interestingly, these variants are active in many tissues (Supp table h4_annotated.ods).
%
%The variant rs11236797 and the two others in locus 11q13.5 is mainly autoimmune.
%All three regulate the gene EMSY-DT and are active induced pluripotent cells, LCDs and skin.
%
%Another pleiotropic variants rs301806
%rs301807
%rs301802
%rs301817
%rs159963
%rs301816
%in the cytoband 1p36.23 that are involved in allergy, cardiovascular and hypertension.
%This variants control the gene RERE, which is a nuclear receptor corregulator and is involved in neurodevelopmental disorders.\\
